Anomalous waves triggered by abrupt depth changes: laboratory experiments and truncated KdV statistical mechanics

### 2019-11-13 paper2_JNLS/paper.tex
NM

Rearranging the TKdV system \eqref{TKdV} gives
\begin{align}
\label{TKdV2}
&\pd{\uL}{t} = -  \frac{1}{2} C_3 \drat^{-3/2} \, \pd{}{x} \Proj (\uL)^2 - C_2 \drat^{1/2} \, \frac{\partial^3 \uL}{\partial x^3} = \RHS
%\qquad \text{for } x \in [-\pi,\pi]
\end{align}	
where we have represented the right-hand side by $\RHS$.
Our numerical scheme first separates the Hamiltonian into linear and nonlinear components. Then a pseudo-spectral scheme is used for spatial discretization with de-aliasing applied to the nonlinear term $\Proj (\uL)^2$ according to the standard 2/3-rule. That is, we first pad $\uhat_k$ with zeros for $\Lambda < \abs{k} \le 3 \Lambda$ \nick{3 or 3/2???}, transform to physical space and square to obtain $\uL^2$ on a fine grid, then transform back to spectral space and truncate to obtain the Fourier coefficients of $\Proj (\uL)^2$
For simplicity, we denote these Fourier coefficients $\usqdah_k$. That is,
\begin{equation}
\Proj (\uL)^2 = \sum_{\abs{k} \le \Lambda} \usqdah_k e^{i k x}
\end{equation}

@@

%The symplectic numerical integrators discretize the Hamiltonian and the Poisson structures separated and then form the resulting scheme for conserving the Hamiltonian.
%In the numerical discretization, first we split the Hamiltonian into separated linear and nonlinear dynamics.

Therefore the ODE system after spatial discretization can be written for the spectral vector $\mathbf{u}=\left\{ \hat{u}_{k}\right\}_{k=0}^{\Lambda}$ and $\mathbf{F}=\left\{ \RHSh_{k} \right\}_{k=0}^{\Lambda}$
\begin{equation}
\dot{\hat{u}}_{k} = \RHSh_k \left[u\right] = C_{2}\mathcal{D}^{1/2}ik^{3}\hat{u}_{k}-\frac{1}{2}C_{3}\mathcal{D}^{-3/2}ik\widehat{\left(u^{2}\right)}_{k}.
\end{equation}
Different scales are mixed due to the nonlinear term $u^{2}$. The linear term including third derivative might become stiff when the total wavenumbers become large.

The symplectic scheme
used here for the time integration of the TKdV equation is the 4th-order midpoint method \cite{mclachlan1993symplectic}.
Let $\mathbf{u}^{n}$ be the solution at time $t_{n}$ and $\mathbf{u}^{n+1}$
be the solution at next time step $t_{n+1}=t_{n}+\Delta t$, and $\mathbf{F}$ the
discretized operator for the spectral modes. The midpoint method has
two intermediate stages and the auxiliary vectors $\mathbf{y}_{1},\mathbf{y}_{2}$ so that
\begin{align}
&\mathbf{y}_{1}-\mathbf{u}^{n} = w_{1}\Delta t\mathbf{F}\left(\frac{1}{2}\left[\mathbf{y}_{1}+\mathbf{u}^{n}\right]\right),\\
&\mathbf{y}_{2}-\mathbf{y}_{1} = w_{2}\Delta t\mathbf{F}\left(\frac{1}{2}\left[\mathbf{y}_{2}+\mathbf{y}_{1}\right]\right),\\
&\mathbf{u}^{n+1}-\mathbf{y}_{2} = w_{3}\Delta t\mathbf{F}\left(\frac{1}{2}\left[\mathbf{u}^{n+1}+\mathbf{y}_{2}\right]\right),
\end{align}
with the time increments $w_{1}=\left(2+2^{1/3}+2^{-1/3}\right)/3$,
$w_{2}=1-2w_{1}$, and $w_{3}=w_{1}$. Then the PDFs can be counted
from the solution $u_{\Lambda}\left(t\right)$, and the statistical
expectations can be calculated by averaging along the trajectory
\begin{equation}
\left\langle F\left(u_{\Lambda}\right)\right\rangle _{\mu}=\frac{1}{T}\int_{t_{0}}^{t_{0}+T}F\left(u_{\Lambda}\left(t\right)\right)dt,
\end{equation}
given ergodicity in the solution with $u_{\Lambda}\left(t_{0}\right)$
already reaching the stationary state.

@@

>>>>

Rearranging the TKdV system \eqref{TKdV} gives
\begin{align}
\label{TKdV2}
&\pd{\uL}{t} = -  \frac{1}{2} C_3 \drat^{-3/2} \, \pd{}{x} \Proj (\uL)^2 - C_2 \drat^{1/2} \, \frac{\partial^3 \uL}{\partial x^3} = \RHS
%\qquad \text{for } x \in [-\pi,\pi]
\end{align}	
where we have represented the right-hand side by $\RHS$.
Our numerical scheme first separates the Hamiltonian into linear and nonlinear components. Then a pseudo-spectral scheme is used for spatial discretization with de-aliasing applied to the nonlinear term $\Proj (\uL)^2$ according to the standard 2/3-rule. That is, we first pad $\uhat_k$ with zeros for $\Lambda < \abs{k} \le 3\Lambda/2$, transform to physical space and square to obtain $\uL^2$ on a fine grid, then transform back to spectral space and truncate to obtain the Fourier coefficients of $\Proj (\uL)^2$
For simplicity, we denote these Fourier coefficients $\usqdah_k$. That is,
\begin{equation}
\Proj (\uL)^2 = \sum_{\abs{k} \le \Lambda} \usqdah_k e^{i k x}
\end{equation}
With this spectral discretization, the ODE system \eqref{TKdV2} can be written for the spectral vector $\mathbf{u}=\left\{ \hat{u}_{k}\right\}_{k=0}^{\Lambda}$ and $\mathbf{F}=\left\{ \RHSh_{k} \right\}_{k=0}^{\Lambda}$ as
\begin{equation}
\label{TKdVDisc}
\td{}{t} \uhat_k =  -\frac{1}{2} C_3 \drat^{-3/2} \, ik \usqdah_k + C_2 \drat^{1/2} \, ik^{3} \uhat_k = \RHSh_k
\end{equation}
Due to the nonlinearity, this system mixes scales during evolution. We note the third-order linear term may become stiff for large $\Lambda$.

For time integration of \eqref{TKdVDisc}, we employ a 4th-order midpoint symplectic scheme \cite{mclachlan1993symplectic}. Let $\mathbf{u}^{n}$ be the solution at time $t_{n}$ and $\mathbf{u}^{n+1}$
be the solution at next time step $t_{n+1}=t_{n}+\Delta t$, and $\mathbf{F}$ the
discretized operator for the spectral modes. The midpoint method has
two intermediate stages and the auxiliary vectors $\mathbf{y}_{1},\mathbf{y}_{2}$ so that
\begin{align}
&\mathbf{y}_{1}-\mathbf{u}^{n} = w_{1}\Delta t\mathbf{F}\left(\frac{1}{2}\left[\mathbf{y}_{1}+\mathbf{u}^{n}\right]\right),\\
&\mathbf{y}_{2}-\mathbf{y}_{1} = w_{2}\Delta t\mathbf{F}\left(\frac{1}{2}\left[\mathbf{y}_{2}+\mathbf{y}_{1}\right]\right),\\
&\mathbf{u}^{n+1}-\mathbf{y}_{2} = w_{3}\Delta t\mathbf{F}\left(\frac{1}{2}\left[\mathbf{u}^{n+1}+\mathbf{y}_{2}\right]\right),
\end{align}
with the time increments $w_{1}=\left(2+2^{1/3}+2^{-1/3}\right)/3$,
$w_{2}=1-2w_{1}$, and $w_{3}=w_{1}$.

@@

Then the PDFs can be counted from the solution $u_{\Lambda}\left(t\right)$, and the statistical expectations can be calculated by averaging along the trajectory
\begin{equation}
\left\langle F\left(u_{\Lambda}\right)\right\rangle _{\mu}=\frac{1}{T}\int_{t_{0}}^{t_{0}+T}F\left(u_{\Lambda}\left(t\right)\right)dt,
\end{equation}
given ergodicity in the solution with $u_{\Lambda}\left(t_{0}\right)$ already reaching the stationary state.

@@

%The symplectic numerical integrators discretize the Hamiltonian and the Poisson structures separated and then form the resulting scheme for conserving the Hamiltonian.
